Measurement linkage using Markov Link Intervals

### 2019-05-13 paper/paper.tex
More SCIENCE

It is particularly challenging to estimate the link when multiple measurements of the same specimen are unavailable.  For example, it is currently often difficult to observe both the RNA expression and morphological properties of a single cell.  Rather, we typically have two datasets: one contains the transcriptomes of one group of cells and the other contains the morphology of another group of cells from the same population.   However, the datasets are not completely disconnected: there is some information which is collected for every specimen in both datasets.  For example, it may be possible to read optogenetic marker information for every cell in both datasets.  The task is to leverage this information to uncover the link.

@@

The key challenge in unpaired measurement linkage is to identify good assumptions or inductive biases; without any additional assumptions the problem is always ill-posed.  An overview of different possible assumptions can be found in Fan \emph{et al} \cite{FanYan2017}.

In this paper, we seek to understand unpaired measurement linkage in the discrete setting under the assumption that $L$ and $Y$ are conditionally independent given $X$.  This assumption is inspired by cell biology, and holds naturally whenever $L$ and $Y$ can be understood as independent noisy measurements of $X$.  For example, in Section \ref{sec:allen} we will see an example where optogenetic markers ($L$) and a low-fidelity transcriptomic record ($Y$) act as independent noisy indicators about cell-type ($X$).  Many other examples can be found in Section \ref{sec:conclusions}.  Note that the side-information $L$ bears a formal similarity to an instrumental variable, but its function is different; instrumental variables resolve causal effects in the presence of unobserved confounders, whereas these variables permit joint inference when only marginal observations are available.

We investigate the identifiability issues and develop estimators which are asymptotically guaranteed to bound the true joint distribution.  The key idea is to make a detailed study of the ``observational equivalence classes'' \cite{paulino1994identifiability,hauser2012characterization,tong1991indeterminacy}.  Each class corresponds to a set of parameters that give rise to the same distribution on the data we observe.  We show that estimating these sets is actually impossible, but a carefully-chosen dilation of these sets asymptotically contains the true model parameters and also concentrates on the observable distributions.  We also design a simple optimization algorithm to find extremal points in these dilated sets.  These extremal points yield bounds for the parameters which we dub ``Markov Link Intervals,'' because they use the Markov conditional independence assumption to bound the parameter that links $X$ to $Y$.   Finally, we interrogate the effectiveness of these intervals using simulations and real-world data.

@@

@@

>>>>

It is particularly challenging to estimate the link when multiple measurements of the same specimen are unavailable.  For example, it is currently often difficult to observe both the RNA expression and morphological properties of a single cell.  Rather, we typically have two datasets: one contains the transcriptomes of one group of cells and the other contains the morphology of another group of cells from the same population.   However, the datasets are not completely disconnected: there is some information which is collected for every specimen in both datasets.  For example, it may be possible to read optogenetic marker information for every cell in both datasets.  The task is to leverage this common information to uncover the link.  Figure \ref{fig:bigpicture} sketches an example of this difficult situation, which is common in biological problems \cite{gouwens2018classification}.

@@

The key challenge in unpaired measurement linkage is to identify good assumptions or inductive biases; without any additional assumptions the problem is always ill-posed.  An overview of many typical assumptions can be found in Fan \emph{et al} \cite{FanYan2017}.

In this paper, we investigate a new assumption: that $L$ and $Y$ are conditionally independent given $X$.  We try to understand theory and practice for unpaired measurement linkage in the discrete setting under this assumption.  This assumption is inspired by cell biology and holds naturally whenever $L$ and $Y$ can be understood as independent noisy measurements of $X$.  For example, in Section \ref{sec:allen} we will see an example where optogenetic markers ($L$) and a low-fidelity transcriptomic measurement ($Y$) act as independent noisy indicators about cell-type ($X$).  Many other examples can be found in Section \ref{sec:conclusions}.  Note that the side-information $L$ bears a formal similarity to an instrumental variable, but its function is different; instrumental variables resolve causal effects in the presence of unobserved confounders, whereas in our case $L$ permits estimation of joint distributions even though joint observations are unavailable.

\begin{figure}\fbox{\begin{minipage}{\textwidth}
\begin{tabular}{cc}
\includegraphics[width=.5\textwidth]{bigpicture.png} 
& \includegraphics[width=.5\textwidth]{graphicalmodel.png} \\
%
\begin{minipage}{.4\textwidth}(a) Setup: each specimen can be measured with one of two procedures,  but can't be measured with both. How do we study the relationship  between the procedures?\end{minipage}
& \begin{minipage}{.3\textwidth}(b) Key modeling assumption: \\ $X$ separates $Y$ from $L$.\end{minipage} 
\end{tabular}
\caption{{\bf Markov Link Intervals use a conditional independence assumption to estimate joint distributions from marginal observations.}  In subfigure (a) we have a population of neurons.  We select half of the neurons and determine what type of neuron they are using one kind of measurement ($X$).  We use a noisier measurement for the second half of neurons and use it to guess their cell-types ($Y$).  To reconcile the two resulting notions of cell-type, we represent them as two random variables on a common probability space.  Markov Link Intervals (defined formally in Section \ref{sec:mlidef}) provides a way to estimate the joint distribution of these two random variables.  In subfigure (b) we present a graphical model diagram articulating the assumption upon which our analysis is based.  For each neuron in each group, this diagram considers the cell-type ($X$), the noisy cell-type guess ($Y$), and an optogentic marker ($L$) which can be observed for every cell.  We assume that the optogenetic marker and the noisy estimate become independent if we condition upon the cell-type; this assumption allows us to estimate the joint distribution.  \label{fig:bigpicture}}
\end{minipage}
}
\end{figure}

We investigate the identifiability issues and develop estimators which are asymptotically guaranteed to bound the true joint distribution.  The key idea is to make a detailed study of the ``observational equivalence classes'' \cite{paulino1994identifiability,hauser2012characterization,tong1991indeterminacy}.  Each class corresponds to a set of parameters that give rise to the same distribution on the data we observe.  We show that estimating these sets is actually impossible, but a carefully-chosen dilation of these sets asymptotically contains the true model parameters and also concentrates on the observable distributions.  We also design an optimization algorithm to find extremal points in these dilated sets.  These extremal points yield bounds for the parameters which we dub ``Markov Link Intervals,'' because they use the Markov conditional independence assumption to bound the parameter that links $X$ to $Y$.   Finally, we interrogate the effectiveness of these intervals using simulations and real-world data.

@@

\label{sec:mainMLI}

@@

\label{sec:mlidef}
